A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-04 paint-contour-tree.tex
small edits

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$.
This is at most $O(t\log D + N)$, where $D$ is the diameter of the contour tree.
Moreover, it is $O(t\alpha(t)+N)$ for balanced trees, a significant improvement over the previous complexity.

Our algorithm requires numerous ideas: partitioning the contour tree into join and split trees, 
a local growing procedure to iteratively build contour trees, and the use of heavy path 
decompositions for the time complexity analysis. There is a crucial use of a family of binomial heaps to 
maintain priorities, ensuring that any comparison made is between comparable nodes of the contour tree.
We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
is inherent to computing contour trees. 
%Consider the family of instances where our algorithm
%requires $C$ comparisons. Any algorithm computing contour trees on this family requires
%$\Omega(C)$ comparisons in the worst case.

@@

\begin{theorem}
Consider a PL-Morse $f:\MM \mapsto \RR$, where the contour tree $\cC=\cC(\MM)$
has maximum degree $3$. There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}
\begin{proof}
First, lets review the various pieces of our algorithm. 
On a given input simplicial complex, we first we break it into extremum dominant pieces using $\rain(\MM)$ (and in $O(|\MM|)$ time by \Thm{rain-time}).  
Specifically, \Lem{rain-1} proves that the output of $\rain(\MM)$ is a set of extremum dominant pieces, 
$\MM_1, \dots, \MM_k$, and \Clm{rain-reeb} shows that given the contour trees, $\cC(\MM_1), \dots, \cC(\MM_k)$ , 
the full contour tree, $\cC(\MM)$, can be constructed (in $O(|\MM|)$ time).

@@

%In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
%Observe that given a fixed $(f,\MM)$, 
%our algorithm for computing $\jc$ is deterministic and so the maximum path decomposition $P=P(\jc)$ defined by the initial coloring (\Def{initialColoring}) is fixed, 
%allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.
%  
We focus on terrains, so $d=2$.
Consider any valid path decomposition $P$ of a valid join tree (actually, for any rooted binary tree,
there exist a terrain for this is the contour tree).
When we say ``compute the join tree", we require the join tree to
be labeled with the corresponding vertices of the terrain.

@@

\begin{lemma}
\label{lem:pathDecomp}
Fix a valid path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
\end{lemma}

@@

%The main idea is that 
% this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
% whose size is that described in the lemma statement. 
 We describe the basic building block of these terrains, which corresponds to a fixed path $p\in P$.  
 Informally, a \emph{diamond} is an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
 Construct a slightly tilted cycle 
 of length $m$ with the two antipodal points at heights $1$ and $0$. These are called the anchor and trap of the diamond, respectively.
 The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
 Next, create an apex vertex at some appropriately large height, and add an edge each vertex in the cycle.

@@

 There is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
 In particular, there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
 terrains we can build.  Each one of these functions will result in a different (labeled) join tree. All saddles on a given diamond will appear in sorted order in the join tree.
 So, any permutation of the heights on a given diamond corresponds to a permutation of the vertices along a path in $P$.
\end{proof}

@@

\begin{lemma}
\label{lem:cost}
For all $\MM \in \bF_P$, the total number of heap operations performed by $\build(\MM)$ is $\Theta(\sum_{p\in P} |p|\log|p|)$.
\end{lemma}
\begin{proof}
%For any choice of $\MM \in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree equivalent to $\mathcal{P}$.
%Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
The primary ``non-determinism" of the algorithm is the initial painting constructed by $\init(\MM)$. 
We show that regardless of how paint spilling is done, the number of heap operations is bounded as above.

 Consider an arbitrary order of the initial paint spilling over the surface defined by $f$.
 Consider any join on a face of some diamond, which is the anchor point of some connecting child diamond.  
 The join has two up-stars, each of which has exactly one edge. Each edge connects to a maxima
 and must be colored by that maxima.
 Hence, two colors touching this join (according to \Def{initialColoring})
 are the colors of the apexes of the child and parent diamond.  
% 
% Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
% The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
% so the second color is that of the maxima of the child diamond.

@@

\begin{theorem}\label{thm:joinLB}
Consider a rooted tree $T$ and an arbitrary path decomposition $P$ of $t$.
There is a family $\bF_P$ of terrains such that any algebraic decision
tree computing the contour tree (on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ on all these instances.
%
%Consider the set of functions $\bF_C$ that all have the same join tree valid cost $C$.  
%Any height function decision tree that correctly computes
%the join tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
\end{theorem}

\begin{proof}
The proof is a basic entropy argument. Any algebraic decision tree that is correct on all of $\bF_P$
must distinguish all inputs in this family. By Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\Lem{cost} completes the proof.
%
%
% If $C$ is a join tree valid cost then there some $f$ such that $\cost_{\jc}(f)=\sum_{p\in P(f)} |p|\log|p|$.
% By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
% (Specifically, each sorted ordering of the relative height values of the saddles on each diamond yields a function in this class.)
% Hence $\bF\subseteq \bF_C$, and so any height function decision tree for the set $\bF_C$ must distinguish between these possibilities, and therefore must 
% have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\end{proof}

@@

\begin{lemma}
\label{lem:pathDecomp2}
Let $P$ be a rain consistent maximum path decomposition of some contour tree. 
There is a family of terrains, $\bF$, all with the same triangulation, 
such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two functions in $\bF$ define the same contour tree.
\end{lemma}
\begin{proof}
 As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
 vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
 let $r_i$ be the root vertex of this subtree.  
 Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e. the root of $T_i$) is adjacent to a vertex in $P_j$.

@@

 So for each $P_i$ construct a terrain as described in \Lem{pathDecomp}.  Now each $T_i$ is (roughly speaking) a join or a split tree, depending on 
 whether the BFS which produced it was an upward or downward BFS, respectively.  As the construction in \Lem{pathDecomp} was for join trees, for each terrain 
 we constructed for a $P_i$ which came from a split tree, must be flipped upside down.
 Now we must described how to glue the terrains together.

@@

 Just as in \Lem{pathDecomp} we now have one fixed terrain structure, such that each different relative ordering of the heights of the join and split vertices 
 on each diamond produces a surface with a distinct join tree.  The specific bound on the size of $\bF$, defining these distinct join trees, follows by applying the 
 bound from \Lem{pathDecomp} to each $P_i$.
\end{proof}

@@

\Thm{main-lb} follows immediately from an entropy argument, analogous to \Thm{joinLB}.
%
%
%Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
%For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following. 
%
%\begin{theorem}
%Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
%Any height function decision tree that correctly computes
%the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
%\end{theorem}

>>>>

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$ time.
This is at most $O(t\log D + N)$, where $D$ is the diameter of the contour tree.
Moreover, it is $O(t\alpha(t)+N)$ for balanced trees, a significant improvement over the previous complexity.

Our algorithm requires numerous ideas: partitioning the contour tree into join and split trees, 
a local growing procedure to iteratively build contour trees, and the use of heavy path 
decompositions for the time complexity analysis. There is a crucial use of a family of binomial heaps to 
maintain priorities, ensuring that any comparison made is between nodes in the same path in the contour tree.
We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
is inherent to computing contour trees. 
%Consider the family of instances where our algorithm
%requires $C$ comparisons. Any algorithm computing contour trees on this family requires
%$\Omega(C)$ comparisons in the worst case.

@@

\begin{theorem}
Consider a PL-Morse $f:\MM \mapsto \RR$, where the contour tree $\cC=\cC(\MM)$
has maximum degree $3$. There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}
\begin{proof}
First, lets review the various pieces of our algorithm. 
On a given input simplicial complex, we first break it into extremum dominant pieces using $\rain(\MM)$ (and in $O(|\MM|)$ time by \Thm{rain-time}).  
Specifically, \Lem{rain-1} proves that the output of $\rain(\MM)$ is a set of extremum dominant pieces, 
$\MM_1, \dots, \MM_k$, and \Clm{rain-reeb} shows that given the contour trees, $\cC(\MM_1), \dots, \cC(\MM_k)$, 
the full contour tree, $\cC(\MM)$, can be constructed (in $O(|\MM|)$ time).

@@

%In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
%Observe that given a fixed $(f,\MM)$, 
%our algorithm for computing $\jc$ is deterministic and so the maximum path decomposition $P=P(\jc)$ defined by the initial coloring (\Def{initialColoring}) is fixed, 
%allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.
%  
We focus on terrains, so $d=2$.
Consider any path decomposition $P$ of a valid join tree (i.e. any rooted binary tree).
When we say ``compute the join tree", we require the join tree to
be labeled with the corresponding vertices of the terrain.

@@

\begin{lemma}
\label{lem:pathDecomp}
Fix a path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
\end{lemma}

@@

%The main idea is that 
% this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
% whose size is that described in the lemma statement. 
 We describe the basic building block of these terrains, which corresponds to a fixed path $p\in P$.  
 Informally, a \emph{diamond} is an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
 Construct a slightly tilted cycle 
 of length $m$ with the two antipodal points at heights $1$ and $0$. These are called the anchor and trap of the diamond, respectively.
 The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
 Next, create an apex vertex at some appropriately large height, and add an edge to each vertex in the cycle.

@@

 There is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
 In particular, there are $(|p_i|-1)!$ possible orderings of the heights of the saddles on the diamond for $p_i$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
 terrains we can build.  Each one of these functions will result in a different (labeled) join tree. All saddles on a given diamond will appear in sorted order in the join tree.
 So, any permutation of the heights on a given diamond corresponds to a permutation of the vertices along a path in $P$.
\end{proof}

@@

\begin{lemma}
\label{lem:cost}
For all $\MM \in \bF_P$, the total number of heap operations performed by $\build(\MM)$ is $O(\sum_{p\in P} |p|\log|p|)$.
\end{lemma}
\begin{proof}
%For any choice of $\MM \in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree equivalent to $\mathcal{P}$.
%Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
The primary ``non-determinism" of the algorithm is the initial painting constructed by $\init(\MM)$. 
We show that regardless of how paint spilling is done, the number of heap operations is bounded as above.

 Consider an arbitrary order of the initial paint spilling over the surface.
 Consider any join on a face of some diamond, which is the anchor point of some connecting child diamond.  
 The join has two up-stars, each of which has exactly one edge. Each edge connects to a maxima
 and must be colored by that maxima.
 Hence, the two colors touching this join (according to \Def{initialColoring})
 are the colors of the apexes of the child and parent diamond.  
% 
% Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
% The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
% so the second color is that of the maxima of the child diamond.

@@

\begin{theorem}\label{thm:joinLB}
Consider a rooted tree $T$ and an arbitrary path decomposition $P$ of $T$.
There is a family $\bF_P$ of terrains such that any algebraic decision
tree computing the contour tree (on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons on all these instances.
%
%Consider the set of functions $\bF_C$ that all have the same join tree valid cost $C$.  
%Any height function decision tree that correctly computes
%the join tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
\end{theorem}

\begin{proof}
The proof is a basic entropy argument. Any algebraic decision tree that is correct on all of $\bF_P$
must distinguish all inputs in this family. By Stirling's approximation, the depth of this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\Lem{cost} completes the proof.
%
%
% If $C$ is a join tree valid cost then there some $f$ such that $\cost_{\jc}(f)=\sum_{p\in P(f)} |p|\log|p|$.
% By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
% (Specifically, each sorted ordering of the relative height values of the saddles on each diamond yields a function in this class.)
% Hence $\bF\subseteq \bF_C$, and so any height function decision tree for the set $\bF_C$ must distinguish between these possibilities, and therefore must 
% have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\end{proof}

@@

\begin{lemma}
\label{lem:pathDecomp2}
Let $P$ be a rain consistent maximum path decomposition of some contour tree. 
There is a family of terrains, $\bF_P$, all with the same triangulation, 
such that the size of $\bF_P$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same contour tree.
\end{lemma}
\begin{proof}
 As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
 vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
 let $r_i$ be the root vertex of this subtree.  
 Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e. the root of $T_i$) is adjacent to a vertex in $P_j$.

@@

 So for each $P_i$ construct a terrain as described in \Lem{pathDecomp}.  Now each $T_i$ is (roughly speaking) a join or a split tree, depending on 
 whether the BFS which produced it was an upward or downward BFS, respectively.  As the construction in \Lem{pathDecomp} was for join trees, each terrain 
 we constructed for a $P_i$ which came from a split tree, must be flipped upside down.
 Now we must described how to glue the terrains together.

@@

 Just as in \Lem{pathDecomp} we now have one fixed terrain structure, such that each different relative ordering of the heights of the join and split vertices 
 on each diamond produces a surface with a distinct contour tree.  The specific bound on the size of $\bF_P$, defining these distinct contour trees, follows by applying the 
 bound from \Lem{pathDecomp} to each $P_i$.
\end{proof}

@@

\Thm{main-lb} now follows immediately from an entropy argument, analogous to \Thm{joinLB}.
%
%
%Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
%For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following. 
%
%\begin{theorem}
%Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
%Any height function decision tree that correctly computes
%the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
%\end{theorem}
